Manuscript for Foodborne-Outbreak Project

### 2023-04-23 products/manuscript/Manuscript2.qmd

Foodborne pathogens typically cause serious gastrointestinal infections.  On average, 1 in 6 Americans acquire foodborne illness each year, leading to 128,000 hospitalizations, and 3,000 mortality [@centerforfoodsafetyandappliednutrition2022].  Economic loss associated with foodborne illness is estimated to be greater than 50 billion US dollars annually.  Moreover, it can be particularly challenging to trace the food sources and exposure settings that are associated with these illnesses .  Nevertheless, foodborne disease outbreak surveillance enables the identification of crucial agents that are commonly involved in outbreaks.  In the United States, CDC conducts surveillance for foodborne disease outbreaks through the Foodborne Disease Outbreak Surveillance System [@gould_surveillance_2013].  The primary responsibility for the identification, investigation, and reporting of foodborne disease outbreaks caused by various enteric microorganism, and chemical/toxic agents lies with state, local, and territorial public health departments, in accordance with established scientific protocols and procedures [@hoffman_capacity_2005].  Previous studies have identified that restaurants are sites frequently attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020;@sarno2021].  The etiologies leading to disease encompass a wide range of microorganisms, ranging from bacteria such as *Salmonella* spp. and *E. coli*, to viruses (norovirus and rotavirus), to parasites (*Cryptosporidium* spp. and *Cyclospora* spp), as well as toxins like marine toxins or metal toxins [@cdc2022].  To verify pathogens of greatest concern, common settings associated with foodborne illness, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).

@@

Two datasets were adopted for this research. The first dataset, which was used to build statistical models for the study, was uploaded onto Kaggle in 2017 and originated in CDC’s National Outbreak Reporting System (NORS) website, with 12 variables and more than 16,950 observations (CDC, 2017). The second dataset, which was used for evaluating the model performance, was retrieved from the CDC’s NORS website. The second data set covers reports of foodborne and waterborne disease outbreaks, as well as enteric disease outbreaks spread by contact with environmental sources, infected animals or persons, and other means from 2009 to 2020.

@@

The remaining two of the three initial hypotheses were proved incorrect. Instead of *E. coli*, results overall indicated that *Salmonella* spp. were the most common pathogens linked to foodborne disease; these bacteria contribute to the greatest amount of illnesses and hospitalizations in comparison to other disease agents for our data. As a matter of fact, contaminated poultry, eggs, beef, and pork are the most common food vehicles that contribute to *Salmonella* outbreaks (@jackson_outbreak-associated_2013). The result obtained from a previous study conducted in Poland from 2010 to 2015 studying the distribution of  *Salmonella* serovars identified from food, animals, and feedingstuffs is consistent with our findings (@skarzynska_distribution_2017).  The study revealed that *Salmonella* most frequently recovered from poultry samples, and *S*. Typhimurium being the most prevalent in pork products.

Finally, we observed that food poisoning is most frequently linked to restaurants instead of private households. This result is in agreement with a previous study that revealed over 40 % of outbreaks could be traced back to restaurants (@hedberg_systematic_2006). In contrast, some Chinese researchers reported that the most common exposure setting was household, followed by restaurants (@li_surveillance_2020).

\[Something about machine learning here???\] #you can briefly talk about the algorithm or working principle behind random forest, what attributes may make it the best model?

These results offer insight into common disease characteristics across food poisoning cases, which is greatly useful for epidemiologists and other public health professionals to anticipate which areas to investigate first for future outbreaks.

@@

# References
# Appendix: Tables and Plots

@@

![Figure 3. Pathogen types and Hospitalizations.](images/pathogen_type_v_hospitalizations.png)
![Figure 4. Common bacterial etiology causing hospitalizations from 1998-2015.](images/bacterial_etiology_v_hospitalization_frequency.png)
![Figure 5. Top 3 bacterial pathogens and their common food carrier month from 1998-2015.](images/top3_bacterial_etiology_v_food_hospitalization_frequency.png)
![Figure 5. Leading locations that result in hospitalizations ](images/Location_v_hospitalization.png)
![Figure 6. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png)
![Figure 7. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness.png)
![Figure 8. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png)
![Figure 9. Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)

@@

>>>>

Foodborne pathogens typically cause serious gastrointestinal infections. On average, 1 in 6 Americans acquire foodborne illness each year, leading to 128,000 hospitalizations, and 3,000 mortality [@centerforfoodsafetyandappliednutrition2022]. Economic loss associated with foodborne illness is estimated to be greater than 50 billion US dollars annually. Moreover, it can be particularly challenging to trace the food sources and exposure settings that are associated with these illnesses . Nevertheless, foodborne disease outbreak surveillance enables the identification of crucial agents that are commonly involved in outbreaks. In the United States, CDC conducts surveillance for foodborne disease outbreaks through the Foodborne Disease Outbreak Surveillance System [@gould_surveillance_2013]. The primary responsibility for the identification, investigation, and reporting of foodborne disease outbreaks caused by various enteric microorganism, and chemical/toxic agents lies with state, local, and territorial public health departments, in accordance with established scientific protocols and procedures [@hoffman_capacity_2005]. Previous studies have identified that restaurants are sites frequently attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020; @sarno2021]. The etiologies leading to disease encompass a wide range of microorganisms, ranging from bacteria such as *Salmonella* spp. and *E. coli*, to viruses (norovirus and rotavirus), to parasites (*Cryptosporidium* spp. and *Cyclospora* spp), as well as toxins like marine toxins or metal toxins [@cdc2022]. To verify pathogens of greatest concern, common settings associated with foodborne illness, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).

@@

Two datasets were adopted for this research. The first dataset, which was used to build statistical models for the study, was uploaded onto Kaggle in 2017 and originated in CDC's National Outbreak Reporting System (NORS) website, with 12 variables and more than 16,950 observations (CDC, 2017). The second dataset, which was used for evaluating the model performance, was retrieved from the CDC's NORS website. The second data set covers reports of foodborne and waterborne disease outbreaks, as well as enteric disease outbreaks spread by contact with environmental sources, infected animals or persons, and other means from 2009 to 2020.

@@

The remaining two of the three initial hypotheses were proved incorrect. Instead of *E. coli*, results overall indicated that *Salmonella* spp. were the most common pathogens linked to foodborne disease; these bacteria contribute to the greatest amount of illnesses and hospitalizations in comparison to other disease agents for our data. As a matter of fact, contaminated poultry, eggs, beef, and pork are the most common food vehicles that contribute to *Salmonella* outbreaks [@jackson_outbreak-associated_2013]. The result obtained from a previous study conducted in Poland from 2010 to 2015 studying the distribution of *Salmonella* serovars identified from food, animals, and feedingstuffs is consistent with our findings [@skarzynska_distribution_2017]. The study revealed that *Salmonella* most frequently recovered from poultry samples, and *S*. Typhimurium being the most prevalent in pork products.

Finally, we observed that food poisoning is most frequently linked to restaurants instead of private households. This result is in agreement with a previous study that revealed over 40% of outbreaks could be traced back to restaurants [@hedberg_systematic_2006]. In contrast, some Chinese researchers reported that the most common exposure setting was household, followed by restaurants [@li_surveillance_2020].

Lastly, the random forest model outperformed all other models. It is also interesting to note that bagging resulted in a higher RMSE than random forest each time despite using all predictors in its splitting, which should have greatly reduced variance. However, random forest models take a random subset of the 10 predictors for splitting; average variance is thus reduced from high to moderate--meaning that random forest should have performed worse. Additional future investigation is suggested to uncover why.

Overall, these results offer insight into common disease characteristics across food poisoning cases, which is greatly useful for epidemiologists and other public health professionals to anticipate which areas to investigate first for future outbreaks.

@@

# Appendix: Tables and Plots

@@

![Figure 3. Pathogen types and Hospitalizations.](images/pathogen_type_v_hospitalizations.png) ![Figure 4. Common bacterial etiology causing hospitalizations from 1998-2015.](images/bacterial_etiology_v_hospitalization_frequency.png) ![Figure 5. Top 3 bacterial pathogens and their common food carrier month from 1998-2015.](images/top3_bacterial_etiology_v_food_hospitalization_frequency.png) ![Figure 5. Leading locations that result in hospitalizations](images/Location_v_hospitalization.png) ![Figure 6. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png) ![Figure 7. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness.png) ![Figure 8. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png) ![Figure 9. Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)

@@

# References
